\documentclass[a4paper,UStitle,numberwithinsect=false,autoref]{lipics-v2021}
\hideLIPIcs

\usepackage{microtype}
\usepackage{float}
\usepackage[ruled,vlined]{algorithm2e}
% Render \paragraph as an unnumbered bold run-in head (suppress 2.0.0.1-style numbers).
\setcounter{secnumdepth}{2}
% Absorb tiny (sub-3pt) overfulls from long hyphenated compound terms cleanly.
\emergencystretch=1em

\title{UiB-AlgoRythm: A Robust Anytime Heuristic Solver for\\ Rooted Maximum Agreement Forest}
\titlerunning{A Robust Anytime Heuristic Solver for Rooted MAF}

%% ---------------------------------------------------------------------------
%% AUTHORS -- single \author block: all names with superscript affiliation numbers,
%% then the numbered institution list. Sam Urmian is the contact author.
%% ---------------------------------------------------------------------------
\author{%
Juni Weisteen Bjerde\textsuperscript{1},
Wim Van den Broeck\textsuperscript{1},
Krishnan Dehaleesan\textsuperscript{1},
Yash Hiren More\textsuperscript{1},
Jakob R{\o}dal Skaar\textsuperscript{1},
Tom\'a\v{s} Turek\textsuperscript{1},
Sam Urmian\textsuperscript{2}%
}{%
\textsuperscript{1}\,Department of Informatics, University of Bergen, Norway \\
\textsuperscript{2}\,Department of Informatics and Centre for the Science of Learning \& Technology (SLATE), University of Bergen, Norway \\
\textnormal{\footnotesize
\nolinkurl{juni.bjerde@uib.no},
\nolinkurl{wim.broeck@uib.no},
\nolinkurl{krishnan.dehaleesan@uib.no},
\nolinkurl{yash.h.more@uib.no},
\nolinkurl{jakob.r.skaar@student.uib.no},
\nolinkurl{tomas.turek@uib.no},
\nolinkurl{sam.urmian@uib.no}~(contact)}%
}{}{}{}

\authorrunning{J.\,W.\,Bjerde et al.}
\Copyright{Juni Weisteen Bjerde, Wim Van den Broeck, Krishnan Dehaleesan, Yash Hiren More, Jakob R{\o}dal Skaar, Tom\'a\v{s} Turek, and Sam Urmian}

\ccsdesc[500]{Theory of computation~Fixed parameter tractability}
\ccsdesc[300]{Applied computing~Molecular evolution}

\keywords{Maximum agreement forest, phylogenetics, rooted SPR distance, heuristic,
kernelization, anytime algorithm, PACE 2026}

\category{PACE 2026 Heuristic Track solver description}

\begin{document}
\maketitle

\begin{abstract}
We describe \texttt{UiB-AlgoRythm}, our submission to the
PACE~2026 Heuristic Track. The problem is rooted \emph{Maximum Agreement Forest}
(MAF) on two rooted binary phylogenetic trees $T_1,T_2$ over a common leaf set of
size $n$: partition the leaves into the fewest agreement components, each inducing
the same rooted topology in both trees with the components edge-disjoint in each
tree, minimizing the component count $k$ (equivalently the rooted SPR distance
$k-1$). Our solver is a single C++17 file, is anytime, uses at most $8$\,GB, and is
SIGTERM-safe. The design is dominated by one overriding rule of the track:
\emph{a single infeasible or empty output disqualifies the entire submission}.
Feasibility and robustness therefore take precedence over solution quality
everywhere. We combine a lossless Bordewich--Semple subtree kernelization, a
merge-based constructive heuristic with a stack of improvement passes, a
never-regress exact Whidden--Beiko--Zeh branch-and-bound repair on small clean
common clades, and an anytime restart portfolio, all behind an always-on
feasibility gate and an async-signal-safe best-so-far flush.
\end{abstract}

\section{Problem and design constraints}

An \emph{agreement forest} of $T_1$ and $T_2$ is a partition of the leaf set into
components such that (i) each component induces the same rooted topology in both trees
(children unordered, degree-$\le 1$ vertices suppressed) and (ii) the components'
minimal spanning subtrees are edge-disjoint in each tree (node-disjoint, for binary
trees). The objective is to minimize $k$. The input is a PACE \texttt{.nw} instance
(\texttt{\#p 2 $n$}, then two Newick trees on labels $1,\dots,n$); the output is one
Newick component per line, \texttt{;}-terminated, singletons as bare integers.

The track sends \texttt{SIGTERM} at the five-minute mark, then \texttt{SIGKILL}; a
\emph{correct} forest must have been printed by then. Two failure modes are
fatal---an infeasible forest and an \emph{empty} output---the latter typically caused
by an uncatchable \texttt{SIGSEGV} from stack overflow on a deeply nested
(caterpillar/ladder) input. Our engineering targets never triggering either mode.

\section{Algorithm pipeline}

Algorithm~\ref{alg:driver} is the top-level anytime driver; the paragraphs below
elaborate each step.

\begin{algorithm}[H]\footnotesize
\setlength{\intextsep}{5pt plus 2pt minus 2pt}
\DontPrintSemicolon
\SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
\SetKwFunction{Improve}{improveToFixpoint}\SetKwFunction{Sweep}{cladeSweep}
\SetKwFunction{Pub}{publishIfBetter}\SetKwFunction{Publish}{publish}
\SetKwFunction{Shatter}{shatter}\SetKwFunction{Kernelize}{kernelize}
\SetKwFunction{Exact}{exactCorePolish}
\Input{rooted binary trees $T_1,T_2$ on leaves $1,\dots,n$}
\Publish{singletons}\tcp*{valid $k{=}n$; cached and flushed on \texttt{SIGTERM}}
$(T_1,T_2),\pi \leftarrow$ \Kernelize{$T_1,T_2$}\tcp*{Bordewich--Semple; adopt if ${\ge}10\%$ smaller}
$F \leftarrow$ smaller of \Sweep{$T_1$}, \Sweep{$T_2$}\;
$F \leftarrow$ \Improve{$F$};\quad \Pub{$F$}\tcp*{fixpoint of local moves}
$F \leftarrow$ \Exact{$F$};\quad \Pub{$F$}\tcp*{never-regress}
\ForEach{seed $S$ in portfolio}{\Pub{\Improve{$S$}}}
\While{\textnormal{not expired} \textbf{and} $k(F)>1$}{\Pub{\Improve{\Shatter{best}}}\tcp*{LNS restart}}
\Return best\tcp*{expanded through kernel map $\pi$ at output}
\caption{Anytime rooted-MAF heuristic (single worker thread, $1$\,GiB stack).
\textnormal{\textsc{publishIfBetter}$(F)$ installs $F$ as the new best iff $F$ is feasible
and $k(F)$ strictly improves.}}
\label{alg:driver}
\end{algorithm}

\paragraph{Singleton fallback published first.} The moment $n$ is read from the
header---before any search---the all-singleton forest ($k=n$, always feasible) is
cached, so a \texttt{SIGTERM} at any later instant flushes a valid forest rather than
nothing.

\paragraph{Lossless common-subtree kernelization.} We iteratively collapse every
\emph{common cherry} (two leaves that are siblings in both trees) into a single
super-leaf, cascading to collapse maximal common pendant subtrees. This is the
classic Bordewich--Semple subtree reduction~\cite{BordewichSemple2005}: a common
pendant subtree is intra-component in every MAF, so the reduction is
distance/optimality preserving and any valid forest on the kernel expands to a
valid forest on the original instance. We solve the smaller kernel and expand each
super-leaf back to its subtree at output time. The kernel is adopted only on a
meaningful ($\ge 10\%$) reduction; a barely-reduced kernel of a deep common ladder is
discarded, since solving the full instance is then both faster and better.

\paragraph{Construction: common-clade sweep.} The initial forest is built by a
\emph{common-clade sweep}. Scanning one tree's clades in decreasing size, we keep every
clade that also occurs in the other tree and is edge-disjoint from what is already
kept, so the shared backbone becomes a few large components and only the genuinely
conflicting leaves remain as singletons; run from each tree's viewpoint, we keep the
smaller forest.

\paragraph{Improvement: a fixpoint of complementary local moves.} The forest is then
reduced by a stack of local moves, iterated to a fixpoint: singleton pairing and
packing; windowed \emph{component packing} and \emph{corridor merges} that re-group the
components over a small region of one tree; greedy \emph{merges} of two components whose
union is still a common agreement subtree; and a bounded exact-cover \emph{window
repair} that re-optimizes a small window to its local optimum. On large instances a
\emph{conflict-directed} repair unblocks a merge by splitting the blocking component
when that still lowers the count. Every move tests edge-disjointness against the live
forest before it is applied,
so feasibility is an invariant of the entire stack.

\paragraph{Never-regress exact-core repair.} As a final polish on the converged forest,
small \emph{clean} common clades---clades appearing in both trees that are exactly a
union of whole current components---are re-solved to proven optimum by an exact
rooted-MAF branch-and-bound in the style of Whidden, Beiko, and
Zeh~\cite{WhiddenBeikoZeh2013}. The exact core is node- and wall-time-capped; if it does
not finish, the heuristic result is kept, and a solved core is spliced back only when it
\emph{strictly} reduces the component count, so this step can never worsen the forest. Running it early was found to steal
budget from the more productive search phases, so it is deferred to the end.

\paragraph{Seed portfolio and anytime restart.} Because the greedy stack is
basin-dependent, we restart it from a portfolio of structurally different initial
forests: a maximum agreement subtree packing, greedy and greedy-partition
constructors, a cherry-cut beam, exact common-cluster decompositions, and a pair
portfolio. Each is fed through the same improvement stack and adopted only if it beats
the incumbent. The remaining budget drives a large-neighborhood restart loop: it
shatters a random subset of the best forest's components back into singletons
(alternating gentle and aggressive ratios) and lets the stack re-glue that region a
different way, always keeping the best valid forest seen.

\paragraph{Always-on feasibility gate.} Every candidate forest is re-checked before
it may become the published best: the labels must partition $1,\dots,n$, each
component must induce the same rooted topology in both trees, and the components must
be edge-disjoint in both trees. An infeasible candidate is refused and the previous
valid best is retained. The check is near-linear, comparing oversized near-identical
components by a $128$-bit canonical topology hash rather than the quadratic string form
that would blow up on deep shapes.

\section{Robustness engineering}

The distinctive part of the submission is defensive: correctness must survive
adversarial input shapes and an asynchronous kill.

\paragraph{Deep-input stack safety.} Every recursive routine (Newick parsing and
serialization, restricted-topology construction, kernel expansion) recurses to a
depth that is linear in the tree \emph{depth}, which for a caterpillar is $n$. On the
default $\approx 8$\,MB stack this overflows and \texttt{SIGSEGV}s---silently, with
empty output---at $n$ of the order of $10^5$. We run the entire pipeline (parse,
kernelize, solve, flush) on a single worker thread whose $1$\,GiB stack is set on the
thread attribute, independently of the \texttt{ulimit} soft limit. Only touched pages
are resident, costing no real memory on the shallow common case. The process stays
single-threaded in the PACE sense: the main thread only blocks joining the worker.

\paragraph{Async-signal-safe termination.} The best-so-far forest is kept as a single
pre-serialized, newline-terminated string in a double-buffered cache. The
\texttt{SIGTERM}/\texttt{SIGINT} handler performs no allocation and no locking: it
reads the current buffer pointer and length, writes them to file descriptor~$1$ with
a single \texttt{write}, and exits. On normal exit we first ignore \texttt{SIGTERM} and
emit through the same write path, so a late signal can neither interleave with nor
duplicate output.

\section{Correctness and verification}

The exact branch-and-bound is fuzz-verified against a brute-force oracle ($2000{+}$
instances, $n\le 11$) and an independent reference implementation ($n\le 15$) with zero
mismatches; kernelization is optimality-preserving on thousands of small instances and
its leaf expansion is a bijection. Beyond the always-on feasibility gate, every output
is cross-checked by an independent dependency-free checker that shares no code with the
solver. Re-running end to end for this submission: the official example is solved
optimally; across $550$ random instances ($250$ built to trigger kernelization, which
fired on $62$) every output was feasible and matched the brute-force optimum ($0$
infeasible, $0$ suboptimal); and a deep caterpillar of $n{=}120{,}000$ under an
$8$\,MB stack neither crashed nor timed out---identical trees collapse to the exact
$k{=}1$ forest, and a mid-search \texttt{SIGTERM} flushes a valid full partition. The
two disqualifying failure modes are thus both covered.

\section{Interface and reproducibility}

The solver is a single C++17 file with no external dependencies, built with
\texttt{g++} at \texttt{-O3 -std=c++17 -pthread}. It reads one instance on
\texttt{stdin} and writes the forest to \texttt{stdout}; the budget is
\texttt{PACE\_TIME\_LIMIT} seconds (default $298$, a margin below the five-minute
mark). Two toggles, \texttt{PACE\_NOKERNEL} and \texttt{PACE\_NOEXACT}, disable
kernelization and exact-core repair for ablation. All algorithms are an original
implementation; no third-party solver source was copied.

%% -------------------------------------------------------------------------
\begin{thebibliography}{9}
\bibitem{BordewichSemple2005}
M.~Bordewich and C.~Semple.
\newblock On the computational complexity of the rooted subtree prune and regraft
distance.
\newblock \emph{Annals of Combinatorics}, 8(4):409--423, 2005.

\bibitem{WhiddenBeikoZeh2013}
C.~Whidden, R.~G.~Beiko, and N.~Zeh.
\newblock Fixed-parameter algorithms for maximum agreement forests.
\newblock \emph{SIAM Journal on Computing}, 42(4):1431--1466, 2013.
\end{thebibliography}

\end{document}
